Methods to prioritise pop-up active transport infrastructure

### 2020-06-18 article.Rmd

1. 'filtered permeability', e.g. as [shown](https://twitter.com/CatrionaSwanson/status/1258322956595453952) in [@salfordcitycouncil_salford_2020]

@@

```{r load}
library(sf)
library(tidyverse)
library(tmap)
region_name = "West Yorkshire"
source("code/build.R")
if(!file.exists("rsf_leeds.Rds")) {
  u = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rsf_leeds.Rds"
  download.file(u, "rsf_leeds.Rds")
}
r_original = readRDS("rsf_leeds.Rds")
if(!file.exists("rtid.csv")) {
  ur = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rtid.csv"
  download.file(ur, "rtid.csv")
}
rtid = readr::read_csv("rtid.csv")
# uh = "http://media.nhschoices.nhs.uk/data/foi/Hospital.csv"
# download.file(uh, "uh.csv")
h = readr::read_delim("uh-edited.csv", delim = "|")
# nrow(h) # 1221 hospitals
h_clean = h %>% 
  mutate(Longitude = as.numeric(Longitude)) %>% 
  mutate(Latitude = as.numeric(Latitude)) %>% 
  filter(!is.na(Longitude)) %>% 
  filter(!is.na(Latitude))
hsf = h_clean %>% st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
region = ukboundaries::leeds
tmap_mode("plot")
```

```{r parameters}
regexclude = "welling"
min_cycling_potential = 0
min_grouped_cycling_potential = 100
min_grouped_length = 500
city_centre_buffer_radius_small = 5000
city_centre_buffer_radius = 8000
city_centre_buffer_radius_large = 10000
key_destination_buffer_radius = 5000
city_centre = tmaptools::geocode_OSM("leeds", as.sf = TRUE)
city_centre_buffer_small = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_small)
city_centre_buffer = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius)
city_centre_buffer_large = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_large)
h_leeds = hsf[city_centre_buffer, ]
h_leeds_buffer = stplanr::geo_buffer(h_leeds, dist = key_destination_buffer_radius)
city_key_buffer = st_union(city_centre_buffer, st_union(h_leeds_buffer))
```

```{r preprocess}
# remove motorways
r_original$highway_type = r_original$highway
r_original$highway_type = gsub(pattern = "_link", replacement = "", r_original$highway)
highway_table = table(r_original$highway)
highway_rare = highway_table[highway_table < nrow(r_original) / 100]
highway_remove = names(highway_rare)[!grepl(pattern = "motor|living|ped", x = names(highway_rare))]
r_cleaned = r_original %>% 
  filter(!grepl(pattern = "motorway", x = highway)) %>% 
  mutate(highway_type = case_when(
  highway_type %in% highway_remove ~ "other",
  grepl(pattern = "path|track|trunk", highway_type) ~ "other",
  grepl(pattern = "ped|liv", highway_type) ~ "pedestrian/living_street",
  TRUE ~ highway_type
  )) %>% 
  mutate(maxspeed = case_when(
    maxspeed <= 20 ~ "20 mph or less",
    maxspeed > 20 & maxspeed <= 30 ~ "30 mph",
    maxspeed > 30 ~ "40+ mph",
  )) %>% 
  mutate(cycling_potential = as.numeric(pctgov)) %>% 
  select(name, ref, highway, maxspeed, pctgov, width, Existing, length, highway_type, cycling_potential, rtid, idGlobal)
# table(r_cleaned$maxspeed)
rj = inner_join(r_cleaned, rtid) %>% 
  mutate(lanes_f = abs(lanespsvforward) + abs(lanesforward)) %>% 
  mutate(lanes_b = abs(lanespsvbackward) + abs(lanesbackward)) %>% 
  mutate(n_lanes_numeric = lanes_f + lanes_b) %>% 
  mutate(n_lanes = as.character(n_lanes_numeric))
rj$n_lanes_numeric[rj$n_lanes_numeric <= 0] = 1 
rj$n_lanes[rj$n_lanes == "0"] = "1"
rj$n_lanes[rj$n_lanes == "4"] = "4+"
rj$n_lanes[rj$n_lanes == "5"] = "4+"
rj$n_lanes[rj$n_lanes == "6"] = "4+"
cy = r_cleaned %>% filter(highway == "cycleway")
r_central = rj[city_key_buffer, ]
r_main = r_central %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
r_main_region = rj %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
```

Three key datasets were used for the project:

- Estimates of cycling potential at the street segment level from the UK Department for Transport funded Propensity to Cycle Tool (PCT) project [@goodman_scenarios_2019; @lovelace_propensity_2017]
- Data derived from OpenStreetMap, with several new variables added to support cycling infrastructure planning (see www.cyipt.bike for an overview)
- A list of hospital locations from the UK's National Health Service website [www.nhs.uk](https://www.nhs.uk/about-us/nhs-website-datasets/)
<!-- - Data on the location of road traffic casualties from -->

@@

A map showing the spatial distribution of hospitals in the case study city of Leeds is shown in Figure \@ref(fig:hospitals).

@@

```{r hospitals, fig.cap="Overview map of input data, showing the main highway types and location of hospitals in Leeds"}
m1 = r_main_region %>%
  mutate(`Highway type` = highway_type) %>% 
  mutate(`Cycling potential` = case_when(cycling_potential < 100 ~ 100, TRUE ~ cycling_potential)) %>% 
  tm_shape(bbox = st_bbox(city_centre_buffer)) +
  tm_lines(col = "Highway type", palette = c("green", "black", "blue", "grey"),
           lwd = "Cycling potential", scale = 5, lwd.legend = c(100, 200, 500, 1000),
           lwd.legend.labels = c("0 - 100", "100 - 200", "200 - 500", "500+")) +
  tm_shape(h_leeds) + tm_dots(size = 0.5, col = "ParentName", palette = "Dark2", title = "Hospital group") +
  tm_layout(legend.outside = TRUE)
# + tm_text("OrganisationName")
m1
```

@@

<!-- The region of analysis may seem like a basic consideration: most cities have well-defined administrative zones. -->
<!-- In Leeds and many other cities, it makes sense to focus on the region directly surrounding the city centre, in a kind of 'geographical triage' to omit from the analysis pop-up options in the outskirts and focus valuable attention on the routes that are most likely to serve the highest number of people. -->
We set a modifiable parameter `city_centre_buffer_radius` with an initial value of 8 km (5 miles) to geographically subset potential routes. This represents a distance that most people have the physical ability to cycle.
Figure \@ref(fig:gsub) shows the result of subsetting based on physical distance from the centre vs plotting all possible transport network segments within the city boundaries. 
In some cases, regional geographies mean that roads close to an urban centre do not lie within the city boundaries.
To ensure roads that could serve key destinations were included, the parameter `key_destination_buffer_radius` (initially set to 5 km) was used.
<!-- could say more... -->

```{r gsub, fig.cap="Illustration of geographic subsetting based on distance to a central point (Leeds city centre in this case) rather than based on location within somewhat arbitrarily shaped city boundaries. Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
m1 = tm_shape(r_main_region) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders() +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() 
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
tmap_arrange(m1, m2, nrow = 1)
```

@@

```{r levels, fig.height=3, fig.cap="Illustration of the 'group then filter' method to identify roads with spare lanes that meet threshold values for length and cycling potential"}
r_pct_lanes_all = r_central %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  filter(lanes_f > 1 | lanes_b > 1)
# mapview::mapview(r_pct_lanes)

@@

r_filter_before_grouping = rj %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  filter(lanes_f > 1 | lanes_b > 1) %>% 
  filter(cycling_potential > min_grouped_cycling_potential) %>% 
  filter(length > 100)
tmap_mode("plot")
m0 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes_all) + tm_lines() +
  tm_layout(title = "Roads on which there are spare lanes.")
m1 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_filter_before_grouping) + tm_lines() +
  tm_layout(title = "Filter then group:\n(length > 100, cycling_potential > 100)")
m2 = tm_shape(city_key_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes) + tm_lines("graph_group", palette = "Dark2") +
  tm_layout(title = "Group then filter:\n(length > 500, cycling_potential > 100)")
# todo show group membership with colours
ma = tmap_arrange(m0, m1, m2, nrow = 1)
ma
```

@@

r_pct_top_n = r_pct_top %>% 
  group_by(name) %>% 
  slice(which.max(km_day)) %>% 
  filter(name != "") %>% 
  ungroup() %>% 
  top_n(10, wt = km_day) 
# aim: make the labels go on the longest sections
# %>% 
#   st_cast("LINESTRING") %>% 
#   mutate(length_section = as.numeric(st_length(.))) %>% 
#   group_by(name) %>% 
#   slice(which.max(length_section))
# r_pct_top_n$name
# summary(r_pct_grouped$group_length)
```

@@

```{r res, fig.cap="Results, showing road segments with a spare lane (light blue) and road groups with a minium threshold length, 1km in this case (dark blue). The top 10 road groups are labelled."}
tmap_mode("view")
# tm_shape(r_pct_no_overlap) + tm_lines(col = "lightblue", lwd = 4, alpha = 0.6) +
#   tm_shape(r_pct_top) + tm_lines(col = "blue", lwd = 4, alpha = 0.6) +
#   tm_shape(r_pct_top_n) + tm_text("name") +
#   tm_shape(h_leeds) + tm_dots(size = 0.1)
knitr::include_graphics("figures/results-leeds.png")
```

>>>>

1. 'filtered permeability', where measures such as point closures or contraflow cycle lanes are used to design street networks in which routes for cyclists are faster and more direct than routes for drivers, e.g. as [shown](https://twitter.com/TowerHamletsNow/status/1257564043856019458) in [Tower Hamlets]

@@

```{r load}
library(sf)
library(tidyverse)
library(tmap)
region_name = "West Yorkshire"
source("code/build.R")
if(!file.exists("rsf_leeds.Rds")) {
  u = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rsf_leeds.Rds"
  download.file(u, "rsf_leeds.Rds")
}
r_original = readRDS("rsf_leeds.Rds")
if(!file.exists("rtid.csv")) {
  ur = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rtid.csv"
  download.file(ur, "rtid.csv")
}
rtid = readr::read_csv("rtid.csv")
region = ukboundaries::leeds
tmap_mode("plot")
```

```{r parameters}
regexclude = "welling"
min_cycling_potential = 0
min_grouped_cycling_potential = 100
min_grouped_length = 500
city_centre_buffer_radius_small = 5000
city_centre_buffer_radius = 8000
city_centre_buffer_radius_large = 10000
city_centre = tmaptools::geocode_OSM("leeds", as.sf = TRUE)
city_centre_buffer_small = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_small)
city_centre_buffer = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius)
city_centre_buffer_large = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_large)
```

```{r preprocess}
# remove motorways
r_original$highway_type = r_original$highway
r_original$highway_type = gsub(pattern = "_link", replacement = "", r_original$highway)
highway_table = table(r_original$highway)
highway_rare = highway_table[highway_table < nrow(r_original) / 100]
highway_remove = names(highway_rare)[!grepl(pattern = "motor|living|ped", x = names(highway_rare))]
r_cleaned = r_original %>% 
  filter(!grepl(pattern = "motorway", x = highway)) %>% 
  mutate(highway_type = case_when(
  highway_type %in% highway_remove ~ "other",
  grepl(pattern = "path|track|trunk", highway_type) ~ "other",
  grepl(pattern = "ped|liv", highway_type) ~ "pedestrian/living_street",
  TRUE ~ highway_type
  )) %>% 
  mutate(maxspeed = case_when(
    maxspeed <= 20 ~ "20 mph or less",
    maxspeed > 20 & maxspeed <= 30 ~ "30 mph",
    maxspeed > 30 ~ "40+ mph",
  )) %>% 
  mutate(cycling_potential = as.numeric(pctgov)) %>% 
  select(name, ref, highway, maxspeed, pctgov, width, Existing, length, highway_type, cycling_potential, rtid, idGlobal)
# table(r_cleaned$maxspeed)
rj = inner_join(r_cleaned, rtid) %>% 
  mutate(lanes_f = abs(lanespsvforward) + abs(lanesforward)) %>% 
  mutate(lanes_b = abs(lanespsvbackward) + abs(lanesbackward)) %>% 
  mutate(n_lanes_numeric = lanes_f + lanes_b) %>% 
  mutate(n_lanes = as.character(n_lanes_numeric))
rj$n_lanes_numeric[rj$n_lanes_numeric <= 0] = 1 
rj$n_lanes[rj$n_lanes == "0"] = "1"
rj$n_lanes[rj$n_lanes == "4"] = "4+"
rj$n_lanes[rj$n_lanes == "5"] = "4+"
rj$n_lanes[rj$n_lanes == "6"] = "4+"
cy = r_cleaned %>% filter(highway == "cycleway")
r_central = rj[city_centre_buffer, ]
r_main = r_central %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
r_main_region = rj %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
```

Two key datasets were used for the project:

- Estimates of cycling potential at the street segment level from the UK Department for Transport funded Propensity to Cycle Tool (PCT) project [@goodman_scenarios_2019; @lovelace_propensity_2017]
- Data derived from OpenStreetMap, with several new variables added to support cycling infrastructure planning (see www.cyipt.bike for an overview)
<!-- - Data on the location of road traffic casualties from -->

@@

@@

@@

<!-- The region of analysis may seem like a basic consideration: most cities have well-defined administrative zones. -->
<!-- In Leeds and many other cities, it makes sense to focus on the region directly surrounding the city centre, in a kind of 'geographical triage' to omit from the analysis pop-up options in the outskirts and focus valuable attention on the routes that are most likely to serve the highest number of people. -->
We set a modifiable parameter `city_centre_buffer_radius` with an initial value of 8 km (5 miles) to geographically subset potential routes. This represents a distance that most people have the physical ability to cycle.
Figure \@ref(fig:gsub) shows the result of subsetting based on physical distance from the centre vs plotting all possible transport network segments within the city boundaries. 
In some cases, regional geographies mean that roads close to an urban centre do not lie within the city boundaries.
<!-- could say more... -->

```{r gsub, fig.cap="Illustration of geographic subsetting based on distance to a central point (Leeds city centre in this case) rather than based on location within somewhat arbitrarily shaped city boundaries. Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
m1 = tm_shape(r_main_region) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders() +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_scale_bar() 
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer) + tm_borders() +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
tmap_arrange(m1, m2, nrow = 1)
```

@@

```{r levels, fig.height=3, fig.cap="Illustration of the 'group then filter' method to identify roads with spare lanes that meet threshold values for length and cycling potential"}
r_pct_lanes_all = r_central %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  mutate(spare_lane = lanes_f > 1 | lanes_b > 1) %>%
  filter(spare_lane | width >= 10) # need to add in roads >10m width
# mapview::mapview(r_pct_lanes)

@@

r_filter_before_grouping = rj %>% 
  filter(cycling_potential > min_cycling_potential) %>% 
  filter(lanes_f > 1 | lanes_b > 1) %>% 
  filter(cycling_potential > min_grouped_cycling_potential) %>% 
  filter(length > 100)
tmap_mode("plot")
m0 = tm_shape(city_centre_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes_all) + tm_lines() +
  tm_layout(title = "Roads on which there are spare lanes.")
m1 = tm_shape(city_centre_buffer) + tm_borders(col = "grey") +
  tm_shape(r_filter_before_grouping) + tm_lines() +
  tm_layout(title = "Filter then group:\n(length > 100, cycling_potential > 100)")
m2 = tm_shape(city_centre_buffer) + tm_borders(col = "grey") +
  tm_shape(r_pct_lanes) + tm_lines("graph_group", palette = "Dark2") +
  tm_layout(title = "Group then filter:\n(length > 500, cycling_potential > 100)")
# todo show group membership with colours
ma = tmap_arrange(m0, m1, m2, nrow = 1)
ma
```

@@

r_pct_top_n = r_pct_top %>% 
  group_by(name) %>% 
  slice(which.max(cycling_potential)) %>% 
  filter(name != "") %>% 
  ungroup() %>% 
  top_n(10, wt = km_day) 
# aim: make the labels go on the longest sections
# %>% 
#   st_cast("LINESTRING") %>% 
#   mutate(length_section = as.numeric(st_length(.))) %>% 
#   group_by(name) %>% 
#   slice(which.max(length_section))
# r_pct_top_n$name
# summary(r_pct_grouped$group_length)
```

@@

```{r res, fig.cap="Results, showing road segments with a spare lane (light blue) and road groups with a minium threshold length, 1km in this case (dark blue). The top 10 road groups are labelled."}
tmap_mode("view")
# tm_shape(r_pct_no_overlap) + tm_lines(col = "lightblue", lwd = 4, alpha = 0.6) +
#   tm_shape(r_pct_top) + tm_lines(col = "blue", lwd = 4, alpha = 0.6) +
#   tm_shape(r_pct_top_n) + tm_text("name") +
#   tm_shape(h_leeds) + tm_dots(size = 0.1)
knitr::include_graphics("figures/results-leeds.png") # this needs changing
```
